\textbf{ Interpretable Gaussian Processes for Stellar Light Curves }

### 2020-12-04 tex/gitlinks.tex
better inclination notation

\newcommand{\codelink}[1]{\href{https://github.com/rodluger/starry_process/blob/3fb283601c2627f1ce16864ad46a6aad5a4438cc/tex/figures/#1.py}{\codeicon}\,\,}
\newcommand{\animlink}[1]{\href{https://github.com/rodluger/starry_process/blob/3fb283601c2627f1ce16864ad46a6aad5a4438cc/tex/figures/#1.gif}{\animicon}\,\,}
\newcommand{\prooflink}[1]{\href{https://github.com/rodluger/starry_process/blob/3fb283601c2627f1ce16864ad46a6aad5a4438cc/tex/tests/#1.py}{\raisebox{-0.1em}{\input{tests/#1.tex}}}}
\newcommand{\cilink}[1]{\href{https://dev.azure.com/rodluger/starry_process/_build}{#1}}

>>>>

\newcommand{\codelink}[1]{\href{https://github.com/rodluger/starry_process/blob/966f526cbc69956e8beac5c550a653955342aa28/tex/figures/#1.py}{\codeicon}\,\,}
\newcommand{\animlink}[1]{\href{https://github.com/rodluger/starry_process/blob/966f526cbc69956e8beac5c550a653955342aa28/tex/figures/#1.gif}{\animicon}\,\,}
\newcommand{\prooflink}[1]{\href{https://github.com/rodluger/starry_process/blob/966f526cbc69956e8beac5c550a653955342aa28/tex/tests/#1.py}{\raisebox{-0.1em}{\input{tests/#1.tex}}}}
\newcommand{\cilink}[1]{\href{https://dev.azure.com/rodluger/starry_process/_build}{#1}}

### 2020-12-04 tex/ms.tex
better inclination notation

Fortunately, there is a better way to reduce the number of parameters in
the problem: we can explicitly marginalize over the stellar inclination.
That is, we can write the first two moments of our flux GP as
%
\xxx{FIX THE NOTATION IN THIS SECTION}
%
\begin{align}
    \label{eq:einc}
    \mathrm{E} \Big[ \mathbf{f} \, \Big| \, \bar{\pmb{\theta}}_\star, \pmb{\theta}_\bullet \Big]
     & =
    \mathrm{E} \Big[ \mathbf{1} + \pmb{\mathcal{A}} \, \mathbf{y} \, \Big| \, \bar{\pmb{\theta}}_\star, \pmb{\theta}_\bullet \Big]
    \nonumber \\[0.5em]
     & =
    \mathbf{1} + FOO
\end{align}
%
where
%
\begin{align}
    FOO
     & \equiv
    \mathrm{E} \Big[ \pmb{\mathcal{A}} \, \mathbf{y} \, \Big| \, \bar{\pmb{\theta}}_\star, \pmb{\theta}_\bullet \Big]
    \nonumber \\[0.5em]
     & =
    \mathbf{1} +
    \int
    \pmb{\mathcal{A}}(I, \bar{\pmb{\theta}}_\star) \,
    \mathrm{E} \Big[ \mathbf{y} \, \Big| \, \pmb{\theta}_\bullet \Big]
    p(I) \, \mathrm{d}I
\end{align}
%
and
%
\begin{align}
    \label{eq:Einc}
    \mathrm{E} \Big[ \mathbf{f} \, \mathbf{f}^\top \, \Big| \, \bar{\pmb{\theta}}_\star, \pmb{\theta}_\bullet \Big]
     & =
    \mathrm{E} \Big[ \left(\mathbf{1} + \pmb{\mathcal{A}} \, \mathbf{y}\right) \left(\mathbf{1} + \pmb{\mathcal{A}} \, \mathbf{y}\right)^\top  \, \Big| \, \bar{\pmb{\theta}}_\star, \pmb{\theta}_\bullet \Big]
    \nonumber \\[0.5em]
     & =
    \mathbf{1} \, \mathbf{1}^\top +
    2 \sum FOO +
    BAR
\end{align}
%
where
%
\begin{align}
    BAR
     & \equiv
    \int
    \pmb{\mathcal{A}}(I, \bar{\pmb{\theta}}_\star) \,
    \mathrm{E} \Big[ \mathbf{y} \, \mathbf{y}^\top \, \Big| \, \pmb{\theta}_\bullet \Big] \,
    \pmb{\mathcal{A}}^\top(I, \bar{\pmb{\theta}}_\star)
    p(I) \, \mathrm{d}I
    \quad.
\end{align}
%
The expectations inside the integrals are given by
Equations~(\ref{eq:exp_y_sep}) and (\ref{eq:exp_yy_sep}), respectively, and
$\bar{\pmb{\theta}}_\star$ are all the stellar parameters \emph{except}
the inclination, i.e.,
%
\begin{align}
    \bar{\pmb{\theta}}_\star = \left(
    P \,\,\,
    u_1 \,\,\,
    u_2
    \right)^\top
    \quad,
\end{align}
%
where perhaps $P$ is fixed for each star and the limb darkening coefficients
are shared among all stars, as described above.
%
Thus, if we are able to perform the integrals in Equations~(\ref{eq:mu_marg})
and (\ref{eq:cov_marg}), we can \emph{dramatically} reduce the number of
parameters in our ensemble problem.
%
As we show in Appendix~\ref{sec:inclination}, if we assume that stellar
inclinations are distributed isotropically, these integrals
do in fact have closed-form solutions,
since they are (again) integrals over Wigner matrices.
%
We may then replace our expressions for the GP mean and covariance with
%
\begin{align}
    \label{eq:mu_marg}
    \pmb{\mu}(\bar{\pmb{\theta}}_\star, \pmb{\theta}_\bullet)
     & = \mathrm{E} \Big[ \mathbf{f} \, \Big| \, \bar{\pmb{\theta}}_\star, \pmb{\theta}_\bullet \Big]
    \\
    \label{eq:cov_marg}
    \pmb{\Sigma}(\bar{\pmb{\theta}}_\star, \pmb{\theta}_\bullet)
     & = \mathrm{E} \Big[ \mathbf{f} \, \mathbf{f}^\top \, \Big| \, \bar{\pmb{\theta}}_\star, \pmb{\theta}_\bullet \Big] - \pmb{\mu}(\bar{\pmb{\theta}}_\star, \pmb{\theta}_\bullet) \pmb{\mu}^\top(\bar{\pmb{\theta}}_\star, \pmb{\theta}_\bullet)
    \quad.
\end{align}
%
and use these to compute the marginal likelihood instead.

@@

In this section we will compute the first and second moments of the
inclination distribution (Equations~\ref{eq:einc} and \ref{eq:Einc}),
which allows us to compute the mean and covariance of the process
that describes the flux marginalized over all values of the inclination
(Equations~\ref{eq:mu_marg} and \ref{eq:cov_marg}).

@@

\subsection{Speeding up the computation}
\label{sec:speedup}
%
The expressions in the previous section are a bit of a nightmare,
particularly because of the dimensionality of some of the linear
operators involved. The complexity of the expressions is due to
the fact that the second moment of the spherical harmonic vector
projected onto the sky (Equation~\ref{eq:Eyppkkp}) is time-dependent:
it changes as the star rotates. Computing the second moment of the
\emph{flux} requires computing the outer product of this tensor
with itself, leading to expressions like those in
Equation~(\ref{eq:nightmare}). In addition to being cumbersome to
evaluate, the full second moment matrix $\mathbf{E}_I$
(and hence the flux covariance matrix) is costly to compute.
It is helpful that Equation~(\ref{eq:QllpIiip}) does not depend
on any user inputs and thus may be pre-computed, but even still
we require evaluating the four nested sums
in Equation~(\ref{eq:Eyppkkp})
$\mathcal{O}(l_\mathrm{max}^3)$ times
\emph{for each entry} in the $K \times K$ matrix $\mathbf{E}_I$.

Fortunately, the inner two sums in Equation~(\ref{eq:Eyppkkp})
do not depend on user inputs, so those may be pre-computed,
and Equation~(\ref{eq:Eyppkkp}) may be cast as a straightforward
matrix dot product. However, in practice we find it helpful to
take advantage of the phase independence (i.e., stationarity)
of the covariance of our GP, as we did in \S\ref{sec:inc-mom1}:
any two entries $(E_I)_{k,k'}$ and $(E_I)_{j,j'}$ are
the same if $t_{k} - t_{k'} = t_{j} - t_{j'}$. If the data
happen to be evenly sampled, such that the time difference between
adjacent cadences is constant, then we need only compute the
covariance at a total of $K$ points (as opposed to $K^2$), as the
covariance is a circulant matrix which is fully specified by a
single vector of length $K$.

>>>>

Fortunately, there is a better way to reduce the number of parameters in
the problem: we can explicitly marginalize over the stellar inclination.
Let
%
\begin{align}
    \bar{\pmb{\theta}}_\star = \left(
    P \,\,\,
    u_1 \,\,\,
    u_2
    \right)^\top
    \quad,
\end{align}
%
be the vector of all stellar parameters \emph{except}
the inclination, where perhaps $P$ is fixed for each star and the limb darkening coefficients
are shared among all stars, as described above.
Then we can write the first two moments of our flux GP as
%
\begin{align}
    \label{eq:einc}
    \mathrm{E} \Big[ \mathbf{f} \, \Big| \, \bar{\pmb{\theta}}_\star, \pmb{\theta}_\bullet \Big]
     & =
    \mathrm{E} \Big[ \mathbf{1} + \pmb{\mathcal{A}} \, \mathbf{y} \, \Big| \, \bar{\pmb{\theta}}_\star, \pmb{\theta}_\bullet \Big]
    \nonumber \\[0.5em]
     & =
    \mathbf{1} + \mathbf{e}_I
\end{align}
%
where we define the first moment integral
%
\begin{align}
    \label{eq:eI}
    \mathbf{e}_I
     & \equiv
    \mathrm{E} \Big[ \pmb{\mathcal{A}} \, \mathbf{y} \, \Big| \, \bar{\pmb{\theta}}_\star, \pmb{\theta}_\bullet \Big]
    \nonumber \\[0.5em]
     & =
    \int
    \pmb{\mathcal{A}}(I, \bar{\pmb{\theta}}_\star) \,
    \mathrm{E} \Big[ \mathbf{y} \, \Big| \, \pmb{\theta}_\bullet \Big]
    p(I) \, \mathrm{d}I
\end{align}
%
and
%
\begin{align}
    \label{eq:Einc}
    \mathrm{E} \Big[ \mathbf{f} \, \mathbf{f}^\top \, \Big| \, \bar{\pmb{\theta}}_\star, \pmb{\theta}_\bullet \Big]
     & =
    \mathrm{E} \Big[ \left(\mathbf{1} + \pmb{\mathcal{A}} \, \mathbf{y}\right) \left(\mathbf{1} + \pmb{\mathcal{A}} \, \mathbf{y}\right)^\top  \, \Big| \, \bar{\pmb{\theta}}_\star, \pmb{\theta}_\bullet \Big]
    \nonumber \\[0.5em]
     & =
    \mathbf{1} \, \mathbf{1}^\top +
    \mathbf{1} \, \mathbf{e}_I^\top +
    \mathbf{e}_I \, \mathbf{1}^\top +
    \mathbf{E}_I
\end{align}
%
where we define the second moment integral
%
\begin{align}
    \label{eq:EI}
    \mathbf{E}_I
     & \equiv
    \int
    \pmb{\mathcal{A}}(I, \bar{\pmb{\theta}}_\star) \,
    \mathrm{E} \Big[ \mathbf{y} \, \mathbf{y}^\top \, \Big| \, \pmb{\theta}_\bullet \Big] \,
    \pmb{\mathcal{A}}^\top(I, \bar{\pmb{\theta}}_\star)
    p(I) \, \mathrm{d}I
    \quad.
\end{align}
%
The expectations inside the integrals in the expressions for
$\mathbf{e}_I$ and $\mathbf{E}_I$
are given by
Equations~(\ref{eq:exp_y_sep}) and (\ref{eq:exp_yy_sep}), respectively.
%
If we are able to perform the integrals in those expressions,
we can \emph{dramatically} reduce the number of
parameters in our ensemble problem.
%
As we show in Appendix~\ref{sec:inclination}, if we assume that stellar
inclinations are distributed isotropically, these integrals
do in fact have closed-form solutions.
%
We may then replace our expressions for the GP mean and covariance with
%
\begin{align}
    \label{eq:mu_marg}
    \pmb{\mu}(\bar{\pmb{\theta}}_\star, \pmb{\theta}_\bullet)
     & = \mathrm{E} \Big[ \mathbf{f} \, \Big| \, \bar{\pmb{\theta}}_\star, \pmb{\theta}_\bullet \Big]
    \nonumber                                                                                                                                                                                                                                       \\
     & = \mathbf{1} + \mathbf{e}_I
    \\[0.5em]
    \label{eq:cov_marg}
    \pmb{\Sigma}(\bar{\pmb{\theta}}_\star, \pmb{\theta}_\bullet)
     & = \mathrm{E} \Big[ \mathbf{f} \, \mathbf{f}^\top \, \Big| \, \bar{\pmb{\theta}}_\star, \pmb{\theta}_\bullet \Big] - \pmb{\mu}(\bar{\pmb{\theta}}_\star, \pmb{\theta}_\bullet) \pmb{\mu}^\top(\bar{\pmb{\theta}}_\star, \pmb{\theta}_\bullet)
    \nonumber
    \\
     & =
    \mathbf{E}_I
    -
    \mathbf{e}_I \,
    \mathbf{e}_I^\top
    \quad.
\end{align}
%
and use these to compute the marginal likelihood instead.

@@

In this section we will compute the first and second moment integrals of the
inclination distribution (Equations~\ref{eq:eI} and \ref{eq:EI}),
which allow us to compute the mean and covariance of the process
that describes the flux marginalized over all values of the inclination
(Equations~\ref{eq:mu_marg} and \ref{eq:cov_marg}).

@@

\subsection{Speeding up the computation}
\label{sec:speedup}
%
The expressions in the previous section are a bit of a nightmare,
particularly because of the dimensionality of some of the linear
operators involved. The complexity of the expressions is due to
the fact that the second moment of the spherical harmonic vector
projected onto the sky (Equation~\ref{eq:Eyppkkp}) is time-dependent:
it changes as the star rotates. Computing the second moment of the
\emph{flux} requires computing the outer product of this tensor
with itself, leading to multi-indexed quantities like those in
Equation~(\ref{eq:nightmare}). In addition to being cumbersome to
evaluate, the full second moment matrix $\mathbf{E}_I$
(and hence the flux covariance matrix) is costly to compute.
It is helpful that Equation~(\ref{eq:QllpIiip}) does not depend
on any user inputs and thus may be pre-computed, but even still
we require evaluating the four nested sums
in Equation~(\ref{eq:Eyppkkp})
$\mathcal{O}(l_\mathrm{max}^3)$ times
\emph{for each entry} in the $K \times K$ matrix $\mathbf{E}_I$.

Fortunately, the inner two sums in Equation~(\ref{eq:Eyppkkp})
do not depend on user inputs, so those may be pre-computed,
and Equation~(\ref{eq:Eyppkkp}) may be cast as a straightforward
matrix dot product. In practice we also find it helpful to
take advantage of the phase independence (i.e., stationarity)
of the covariance of our GP, as we did in \S\ref{sec:inc-mom1}:
any two entries $(E_I)_{k,k'}$ and $(E_I)_{j,j'}$ are
the same if $t_{k} - t_{k'} = t_{j} - t_{j'}$. If the data
happen to be evenly sampled, such that the time difference between
adjacent cadences is constant, then we need only compute the
covariance at a total of $K$ points (as opposed to $K^2$), as the
covariance is a circulant matrix which is fully specified by a
single vector of length $K$.
